An elaborate data set on human gait and the effect of mechanical perturbations

### 2014-12-14 paper.tex
Proofed processed data section.

The acceleration of the treadmill was measured during each trial by four
ADXL330 accelerometers placed at the four corners of the machine. These
accelerometers were intended to provide information for inertial compensation
purposes when the treadmill moved laterally, but are extraneous for trials
greater than 8 due to the treadmill being stabilized.
%
\begin{figure*}
  \centering
  \includegraphics{figures/treadmill.png}
  \caption{The treadmill with coordinate system, cameras (circled in orange),
    projection screen, and safety rope. The direction of travel is in the $-z$
    direction.}
  \label{fig:treadmill}
\end{figure*}

@@

\section*{Processed Data}

We developed a toolkit for data processing, GaitAnalysisToolKit, for common
gait computations and provide an example processed trial to show the character
of the data. The tool was developed in Python, is dependent on the SciPy Stack,
and provides two main classes: one to do basic gait data cleaning,
\verb|DFlowData|, and a second to compute common gait variables of interest,
\verb|GaitData|.

The \verb|DFlowData| class collects and stores all the raw data presented in
the previous section and applies several ``cleaning'' operations to transform
the data into a usable form. The cleaning process follows these basic steps:

\begin{enumerate}
  \item Loads the meta data file into a Python dictionary.
  \item Loads the D-Flow mocap module TSV file into Pandas \verb|DataFrame|.
  \item Relabels the column headers to more meaningful names if this is
    specified in the meta data.
  \item Optionally identifies the missing values in the mocap marker data and
    replaces them with \verb|numpy.nan|.
  \item Optionally interpolates the missing marker values and replaces them
    with interpolated estimates using a variety of interpolation methods.
  \item Loads the D-Flow record module TSV file into a Pandas \verb|DataFrame|.
  \item Extracts the events and creates a dictionary mapping the event names in
    the meta data to the events detected in the record module file.
  \item Compensates the ground reaction loads based on whether the meta data
    indicates there was treadmill motion.
  \item Merges the data from the mocap module and record module into one data
    frame at the maximum common constant sample rate.
\end{enumerate}

Once the data is cleaned there are two methods that allow you to extract the
cleaned data:

\begin{itemize}
  \item Extract sections of the data bounded by the events recorded in the
    \verb|record-XXX.txt| file.
  \item Save the cleaned data to disk.
\end{itemize}

An example of the API is given in Listing~\ref{lis:dflowdata}.

\begin{listing}
  \begin{minted}[gobble=4]{python}
    >>> from gaitanalysis.motek import DFlowData
    >>> data = DFlowData('mocap-020.txt', 'record-020.txt', 'meta-020.yml')
    >>> mass = data.meta['subject']['mass']
    >>> data.clean_data()
    >>> event_df = dflow_data.extract_processed_data(event='Longitudinal Perturbation')
  \end{minted}
  \cprotect\caption{Python interpreter session showing how one could load a
    trial into memory, extract the subject's mass from the meta data, run the
    data cleaning process, and finally extract a Pandas \verb|DataFrame|
    containing all of the time histories for a specific event in the trial.}
  \label{lis:dflowdata}
\end{listing}

The \verb|GaitData| class is then used to compute things such as gait landmarks
(toe off and heel strike times), basic 2D kinematics and inverse dynamics, and to store the
data into a Pandas \verb|Panel| with each gait cycle on the main axis at a
specified sample rate. This data is also saved to disk in HDF5 format.

\begin{listing*}
  \begin{minted}[gobble=4]{python}
    >>> from gaitanalysis.gait import GaitData
    >>> gdata = GaitData(event_df)
    >>> gdata.inverse_dynamics_2d(left_markers, right_markers, left_loads,
    ...                           right_loads, mass, 6.0)
    >>> gdata.grf_landmarks('Right Fy', 'Left Fy', threshhold=20.0)
    >>> gdata.split_at('right')
    >>> gdata.plot_gait_cycles('Left Hip Joint Torque', mean=True)
    >>> gdata.save('gait-data.h5')
  \end{minted}
  \cprotect\caption{Python interpreter session showing how one could compute
    the inverse dynamics (joint angles and torques), identify the gait
    landmarks (e.g. heel strikes), split the data with respect to the gait
    landmarks in a Pandas \verb|Panel|, plot the mean and standard deviation of
    one time history with respect to the gait cycles, and save the data to disk.}
  \label{lis:dflowdata}
\end{listing*}

A similar work flow was used to produce Figure~\ref{fig:angle-torque-comparison}
which compares the mean and standard deviation of saggital plane joint angles
and torques from the perturbed gait cycles and the unperturbed gait cycles
computed from trial 20 for 120 gait cycles. The variation in the dynamics is
larger for the perturbed gait cycles.

\begin{figure*}
  \centering
  \includegraphics{figures/unperturbed-perturbed-comparison.pdf}
  \cprotect\caption{Mean (right: solid, left: dashed) and $3\sigma$ (shaded)
    joint angles and torques from both unperturbed (purple) and perturbed
    (blue) gait cycles from trial 20. Produced by \verb|unperturbed_perturbed_comparison.py|.}
  \label{fig:angle-torque-comparison}
\end{figure*}

Figure~\ref{fig:gait-cycle-stats-comparison} compares the distribution in a few
statistics of the unperturbed and perturbed gait cycles.

\begin{figure*}
  \centering
  \includegraphics{figures/unperturbed-perturbed-boxplot-comparison.pdf}
  \cprotect\caption{Box plots of the average belt speed, stride frequency, and
    stride length which compare 120 gait cycles for the unperturbed (purple) and
    perturbed (blue). Produced by \verb|unperturbed_perturbed_comparison.py|.}
  \label{fig:gait-cycle-stats-comparison}
\end{figure*}

>>>>

The acceleration of the treadmill was measured during each trial by four
ADXL330 accelerometers placed at the four corners of the machine. These
accelerometers were intended to provide information for inertial compensation
purposes when the treadmill moved laterally, but are extraneous for trials
greater than 8 due to the treadmill being stabilized.
%
\begin{figure}
  \centering
  \includegraphics{figures/treadmill.png}
  \caption{The treadmill with coordinate system, cameras (circled in orange),
    projection screen, and safety rope. The direction of travel is in the $-z$
    direction.}
  \label{fig:treadmill}
\end{figure}

@@

\section*{Processed Data}
%
We developed a toolkit for data processing,
GaitAnalysisToolKit~\cite{GATK2014}, for common gait computations and provide
an example processed trial to present the nature of the data. The tool was
developed in Python, is dependent on the SciPy Stack and Octave, and provides
two main classes: one to do basic gait data cleaning from D-Flow's output
files, \verb|DFlowData|, and a second to compute common gait variables of
interest, \verb|GaitData|.

The \verb|DFlowData| class collects and stores all the raw data presented in
the previous section and applies several ``cleaning'' operations to transform
the data into a usable form. The cleaning process follows these steps:
%
\begin{enumerate}
  \item Load the meta data file into a Python dictionary.
  \item Load the D-Flow mocap module TSV file into Pandas \verb|DataFrame|.
  \item Relabel the column headers to more meaningful names if this is
    specified in the meta data.
  \item Optionally identify the missing values in the mocap marker data and
    replace them with \verb|numpy.nan|.
  \item Optionally interpolate the missing marker values and replaces them
    with interpolated estimates using a variety of interpolation methods.
  \item Load the D-Flow record module TSV file into a Pandas \verb|DataFrame|.
  \item Extract the events and create a dictionary mapping the event names in
    the meta data to the events detected in the record module file.
  \item Interially compensate the ground reaction loads based on whether the
    meta data indicates there was treadmill motion.
  \item Merge the data from the mocap module and record module into one data
    frame at the maximum common constant sample rate.
\end{enumerate}

Once the data is cleaned there are two methods that allow you to extract the
cleaned data: either extract sections of the data bounded by the events
recorded in the \verb|record-xxx.txt| file or save the cleaned data to disk.
These operations are available as a command line application and as an
application programming interface (API) in Python. An example of the
\verb|DFlowData| API in use is provided in Listing~\ref{lis:dflowdata}.
%
\begin{listing}
  \begin{minted}[gobble=4]{python}
    >>> from gaitanalysis.motek import DFlowData
    >>> data = DFlowData('mocap-020.txt', 'record-020.txt', 'meta-020.yml')
    >>> mass = data.meta['subject']['mass']
    >>> data.clean_data()
    >>> event_df = dflow_data.extract_processed_data(
    ...     event='Longitudinal Perturbation')
  \end{minted}
  \cprotect\caption{Python interpreter session showing how one could load a
    trial into memory, extract the subject's mass from the meta data, run the
    data cleaning process, and finally extract a Pandas \verb|DataFrame|
    containing all of the time histories for a specific event in the trial.}
  \label{lis:dflowdata}
\end{listing}

The \verb|GaitData| class is then used to compute things such as gait events
(toe off and heel strike times), basic 2D kinematics and inverse dynamics, and
to store the data into a Pandas \verb|Panel| with each gait cycle on the item
axis at a specified sample rate. This object can also be serialized to disk in
HDF5 format. An example of using the Python API is shown in
Listing~\ref{lis:gaitdata}.
%
\begin{listing}
  \begin{minted}[gobble=4]{python}
    >>> from gaitanalysis.gait import GaitData
    >>> gdata = GaitData(event_df)
    >>> gdata.inverse_dynamics_2d(left_markers, right_markers, left_loads,
    ...                           right_loads, mass, 6.0)
    >>> gdata.grf_landmarks('Right Fy', 'Left Fy', threshhold=20.0)
    >>> gdata.split_at('right')
    >>> gdata.plot_gait_cycles('Left Hip Joint Torque', mean=True)
    >>> gdata.save('gait-data.h5')
  \end{minted}
  \cprotect\caption{Python interpreter session showing how one could use the
    \verb|GaitData| class to load in the result of \verb|DFlowData| and compute
    the inverse dynamics (joint angles and torques), identify the gait event
    (e.g. heel strikes), split the data with respect to the gait events in a
    Pandas \verb|Panel|, plot the mean and standard deviation of one time
    history with respect to the gait cycles, and save the data to disk.}
  \label{lis:gaitdata}
\end{listing}

A similar work flow was used to produce Figure~\ref{fig:angle-torque-comparison}
which compares the mean and standard deviation of saggital plane joint angles
and torques from the perturbed gait cycles and the unperturbed gait cycles
computed from trial 20. This gives an idea of the more highly variable dynamics
required to walk while being longitudinally perturbed.
%
\begin{figure}
  \centering
  \includegraphics{figures/unperturbed-perturbed-comparison.pdf}
  \cprotect\caption{Mean (right: solid, left: dashed) and $3\sigma$ (shaded)
    joint angles and torques from both unperturbed (purple) and perturbed
    (blue) gait cycles from trial 20. Produced by
    \verb|src/unperturbed_perturbed_comparison.py|.}
  \label{fig:angle-torque-comparison}
\end{figure}

For more insight into the difference in the unperturbed and perturbed data,
Figure~\ref{fig:gait-cycle-stats-comparison} compares the distribution of a few
gait cycle statistics. One can see that the perturbed strides have a much
larger variation in frequency and length.
%
\begin{figure}
  \centering
  \includegraphics{figures/unperturbed-perturbed-boxplot-comparison.pdf}
  \cprotect\caption{Box plots of the average belt speed, stride frequency, and
    stride length which compare gait cycles for the unperturbed (purple) and
    perturbed (blue). Produced by \verb|src/unperturbed_perturbed_comparison.py|.}
  \label{fig:gait-cycle-stats-comparison}
\end{figure}
